```{r setup, include = FALSE}
rm(list = ls())
```

# R as calculator

R is an interactive environment. You can simply use it as a calculator.

```{r}
2 + 2
3 - 7
2 * 2
4^2   # or 4**2
5 / 4
log(3)
exp(1)
```

In longer equations you can use brackets.

```{r}
sqrt(2) * ((5 - 1/6)^2 - pi/2^(1/3))
```
# Creating variables

You can store results in a variable using the `<-` operator.

```{r}
x <- sqrt(2)
y <- x^2
ls()
# rm(list = ls())   # deletes all user defined objects
```

Variable names cannot start with a number. But you can add numbers after a
letter, e.g., `x1` and `x2`.

# Data types

## Vectors

Vectors can be created with the concatenate (or combine) function `c()`.

```{r}
x <- c(6, 3, 2, 8)
x
```

You can use mathematical operations element wise on a vector.

```{r}
mean(x)
x + 2
sqrt(x)
y <- c(2, 4)
x + y   # shorter vector is "recycled" or "broadcasted"
x + rep(y, 2)
```

### Creating vectors

Vectors can be created in different ways. All objects within one vector
must be of the same type. If they are not, R automatically converts all
elements to the most general type.

Often used types of vectors are

* character vectors with their elements in quotation marks
* logical vectors with values `TRUE`, `FALSE` and `NA` (missing values
  indicated by `NA`)

You can check the type with `class()` or `typeof()`.

```{r, fig.dim = c(3.5, 4), fig.align = "center"}
c("a", "b", "c", 4, 7.7)
seq(1, 2, .1)
seq(1, 2, length.out = 10)
seq(from = 1, to = 5, by = 1)
seq(1, 4, along.with = x)
1:10
-5:5
rep(1:5, 2)
rep(c("a", "b", "c"), each = 3)
rep(c("a", "b", "c"), c(2, 4, 6))
table(rep(c("a", "b", "c"), c(2, 4, 6)))
rep(1:4, each = 2, times = 3)         # length 24, 3 complete replications
rep(rep(1:4, each = 2), 3)            # more complicated way
paste("subj", 1:10, sep = "_")
weight <- c(60, 72, 57, 90, 95, 72)
height <- c(1.75, 1.80, 1.65, 1.90, 1.74, 1.91)
plot(height, weight, ylim = c(45, 100))

# logical vectors
c(TRUE, FALSE, FALSE, TRUE)
weight > 60
sum(weight > 60)
mean(weight > 60)
```

Factors are data structures for categorical variables, such as diagnosis,
socio-economic status, sex, etc.

```{r}
ses <- factor(c("low", "inter", "high"))
ses
ses2 <- factor(ses, levels = c("low", "inter", "high"))
ses2
factor(rep(1:2, each = 10), labels = c("on", "of"))
# sort factor levels
l <- paste(c("subj", "cond"), rep(1:10, each = 2), sep = "_")
f <- factor(l, levels = l)
f
```

## Matrices

R offers multiple functionalities to work with matrices.

```{r}
A <- matrix(1:12, nrow = 3, ncol = 4, byrow = TRUE)
A
rownames(A) <- c("a1", "a2", "a3")
t(A)
diag(A)
diag(6)
diag(c(3, 5, 7))
cbind(a1 = 1:4, a2 = 5:8, a3 = 9:12)
rbind(a1 = 1:4, a2 = 5:8, a3 = 9:12)
5 * A
B <- t(A)
B %*% A
```

Arrays are data structures having more than two dimensions.

```{r}
array(c(A, 2 * A), c(3, 4, 2))
```

## Lists

If it is necessary to store different types of R objects into a single
data structure, we need a list.

```{r}
list1 <- list(w = weight, h = height, s = ses2, A = A)
list1
list1$A
list1[[2]]
```

## Data frames

Data frames are the fundamental data structure in R. Data frames are
lists with the restriction that all list elements (column vectors) have the
same length. The rows in a data frame refer to one unit (observation or
subject).

```{r}
id <- factor(paste("s", 1:6, sep = ""))
dat <- data.frame(id, weight, height)
dat
dat$id
```

Frequently used functions (not only) for data frames are

```{r, fig.dim = c(5, 5), fig.align = "center"}
dim(dat)      # show number of rows and columns
names(dat)    # variable names
plot(dat)     # pairwise plots
str(dat)      # show variables of dat
summary(dat)  # descriptive statistics
```

# Indexing

## Indexing vectors

Elements of a vector can be accessed using `[]` (see `?Extract`).

```{r}
weight[4]
weight[4] <- 92
weight[c(1, 2, 6)]
weight[1:5]
weight[-3]
```

Indices may be logical.

```{r}
weight[weight > 60]                # values greater than 60
weight[weight > 60 & weight < 80]  # between 60 and 80
height[weight > 60 & weight < 80]
```

## Indexing data frames

Data frames have a row and a column index. Omitting one index selects all
rows or columns, respectively.

```{r}
dat[3, 2]               # 3rd row, 2nd column
dat[1:4, ]              # rows 1 to 4, all columns
dat[, 3]                # all rows, 3rd column

dat[dat$id == "s2", ]   # all observations of s2
dat[dat$weight > 60, ]  # all observations above 60kg
```

# Calling functions

In R everything is a function. Functions consist of

* a name
* a pair of brackets
* the arguments (none, one, or more)
* a return value (visible, invisible, NULL). 

Functions can be called either with positional or with keyword matching.

```{r, fig.show = "hide"}
# Keyword matching
plot(y = weight, x = height, pch = 16, col = "blue")

# Positional matching
plot(height, weight)
# see ?plot.default

# Mixed
plot(height, weight, pch = 16, col = "blue")
```
Another way to call functions is to use pipes. Pipes are sometimes useful
to not store intermediate objects. Sometimes they are better "readable".

```{r, fig.show = "hide"}
dat |> plot(weight ~ height, data = _)
```


## Useful functions

```{r}
set.seed(1124)      # set seed for reproducibility

dat2 <- data.frame(
  id      = rep(paste0("subj", 1:10), each = 2), 
  cond    = factor(rep(c("c1", "c2"), each = 10)),
  time    = factor(c("t1", "t2")),
  iqscore = rnorm(20, mean = 100, sd = 15)
)
dat2
```

In R, data from different distributions can be randomly generated. See
`?Distributions` for a list of the distributions inbuilt in the stats
package. All of them follow the same schema: there is a root name (`norm`
for the normal distribution); one can get density values for a certain
range with `dnorm()`, quantiles for certain probability values with
`qnorm()`, probability values for certain quantiles with `pnorm()` and
random numbers drawn from the distribution with `rnorm()`.


```{r}
# calculating correlation
t1 <- dat2[dat2$time == "t1", "iqscore"]
t2 <- dat2[dat2$time == "t2", "iqscore"]
cor.test( ~ t1 + t2)
```

### Function `reshape()`

```{r}
dat2[order(dat2$time), ] # sort by time

datw <- reshape(dat2, direction = "wide", idvar = c("id", "cond"),
                timevar = "time")
names(datw)[3:4] <- c("t1", "t2")
cor.test( ~ t1 + t2, datw)

reshape(datw, direction = "long", varying = 3:4, idvar = "id", v.names = "iqscore",
        times = c("t1", "t2"))
reshape(datw, direction = "long", varying = c("t1", "t2"), idvar = "id",
        v.names = "iqscore", times = c("t1", "t2"))

datl <- reshape(datw, direction = "long", varying = c("t1", "t2"), idvar = "id",
                v.names = "iqscore", times = c("t1", "t2"))

datl[order(datl$id), ]

# with more than one independent variable
dat2$math <- rnorm(20)

dat3 <- reshape(dat2, direction = "wide", idvar = c("id", "cond"),
                timevar = "time")

reshape(dat3, direction = "long", varying = list(c(3, 5), c(4, 6)), idvar = "id",
        v.names = c("iqscore", "math"), times = c("t1", "t2"))
```

### Function `aggregate()`

```{r}
aggregate(iqscore ~ cond + time, dat2, mean)
aggregate(cbind(iqscore, math) ~ cond + time, dat2, mean)

aggregate(cbind(iqscore, math) ~ cond + time, dat2, sd)
aggregate(cbind(iqscore, math) ~ cond + time, dat2, var)
aggregate(cbind(iqscore, math) ~ cond + time, dat2, sum)
aggregate(cbind(iqscore, math) ~ cond + time, dat2, length)

# Using a customized function
ms <- function(x) {
  m <- mean(x)
  s <- sd(x)
  n <- length(x)
  se <- s/sqrt(n)
  c(sd = s, mean = m, se = se, n = n)
}
aggregate(cbind(iqscore, math) ~ cond + time, dat2, ms)
```

# Data input/output

The most flexible function to read (text) data into R is `read.table()`.

```{r}
getwd()     # Get working directory
# setwd()   # Set working directory
df <- read.table("data/ourdata.dat", header = TRUE, stringsAsFactors = TRUE)
str(df)
# Fix error
df[df$id == 20, "age"] <- 31
write.table(df, "data/ourdata_corr.dat", row.names = FALSE, quote = FALSE)

# SPSS files
library(foreign)
ds <- read.spss("data/lowbwt.sav", to.data.frame = TRUE)
write.csv2(ds, "data/spss_to_excel.csv")
```

Look up information about `read.table()` with `?read.table`. More
information can be found on the slides.

## Integrated data sets in R

R has plenty of data sets already available. These data sets can be
listed with `data()`. If all available data sets in all installed packages
should be listed use `data(package = .packages(all.available = TRUE))`.

```{r}
rm(list = ls())
data(cars)
ls()
#?cars
```

