# Exercises Day 1

```{r include = FALSE}
# Hide/show R chunks
#knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.show = "hide")
#setwd("C:/Users/nwickelmaier/Nextcloud/Documents/teaching/iwm/R_intro/exercises/")
rm(list=ls())
```

Write an executable and commented R script.

Essential components

* Header
* Set the working directory
* Useful, short, and precise comments

Formal guidelines

* Rows not longer than 80 symbols
* Well-arranged code by using indentions and spaces
* Spaces before and after binary operators (`<-`, `+`, `-`, `~`, `=`, `<`,
  `>`, `<=`, `>=`, `==` etc.)
* *No* space between function name and bracket

After executing the script (with the function `source()`) all computations
should have been executed and all objects should be available in the
working memory.


```{r getdata, include = FALSE}
height <- c(159, 173, 173, 184, 168, 163, 180, 186, 153, 174)
weight <- c(63, 73, 70, 68, 67, 61, 67, 74, 63, 68)
```

1. Create two vectors `height` (in cm) and `weight` (in kg) with the values
   $(`r paste(height, collapse = "~")`)'$ and $(`r paste(weight,
   collapse = "~")`)'$.

   * Compute mean, standard deviation, variance, and correlation for both
     vectors.
   * Create a scatter plot with height on the $x$ axis and weight on the
     $y$ axis. Use meaningful axis labels.
   * Calculate the body mass index for each subject:

$$ \text{BMI} = \frac{\text{weight in kg}}{\text{(height~in~m)}^2}. $$

```{r ex1, fig.height = 4, fig.width = 4, fig.align = "center"}
height <- c(159, 173, 173, 184, 168, 163, 180, 186, 153, 174)
weight <- c(63, 73, 70, 68, 67, 61, 67, 74, 63, 68)

(m_h <- mean(height))
(s_h <- sd(height))
(v_h <- var(height))
(m_w <- mean(weight))
(s_w <- sd(weight))
(v_w <- var(weight))
(kor <- cor(weight, height))
plot(weight ~ height, xlab = "Height", ylab = "Weight")
(bmi <- weight/(height/100)^2)
```

2. Create a column vector $\mathbf{x} = (1~5~8~3~7~2~6)'$. Create another
   column vector $\mathbf{y}$ of the same length containing the odd numbers
   $1, 3, 5, \ldots$ using the function `seq()`. Compute a vector
   $\mathbf{z}$ as linear combination of $\mathbf{x}$ and $\mathbf{y}$:
   $\mathbf{z} = 4\mathbf{x} + 2\mathbf{y}$.

```{r ex2}
x <- c(1, 5, 8, 3, 7, 2, 6)
y <- seq(1, by = 2, along.with = x)
z <- 4 * x + 2 * y
```

3. Combine $\mathbf{x}$, $\mathbf{y}$, and $\mathbf{z}$ into a matrix
   $\mathbf{A}$. Treat $\mathbf{x}$, $\mathbf{y}$, and $\mathbf{z}$ as
   column vectors. Then create a matrix $\mathbf{B}$ with $\mathbf{x}$,
   $\mathbf{y}$, and $\mathbf{z}$ as row vectors. Compute the matrix
   product $\mathbf{B}\mathbf{A}$.

```{r ex3}
A <- cbind(x, y, z)
B <- t(A)
B %*% A
```

4. Create a data frame with two independent variables: *Hand* with
   levels "right" and "left" and *Condition* with levels 1, 2, 3,
   4, and 5. Simulate reaction times for 50 subjects. Assume reaction time
   is normally distributed with $RT \sim N(\mu = 400,\sigma^2 = 625)$. There
   are 10 subjects in each condition. Use functions `str()` and
   `summary()` on your data frame. What does the output tell you?

```{r ex4}
dat <- data.frame(
  hand = factor(rep(c("left","right"), each = 25), levels = c("right","left")),
  cond = factor(rep(1:5, 10)),
  RT   = rnorm(n = 50, mean = 400, sd = sqrt(625))
)
str(dat)
summary(dat)
```

5. Go to
   http://socserv.socsci.mcmaster.ca/jfox/Books/Applied-Regression-3E/ and
   download `Vocabulary.txt` from the data sets there [@Fox2008]. Store
   it in a local folder on your computer. Set that folder as your working
   directory using `setwd()`. Use `read.table()` to load this
   data frame into R's working memory. Hint: With `?read.table` you
   can get documentation for this function. Find out the number of
   observations, number of variables, names of variables, descriptive
   statistics for the dependent variables, and the levels of the factor
   variables.

```{r ex5}
voc <- read.table("data/Vocabulary.txt", header = T, stringsAsFactors = TRUE)
dim(voc)
str(voc)
names(voc)
summary(voc)
```

6. You can use the `$`-operator to extract single variables from the data
   frame "Vocabulary." Extract the variable `vocabulary` and calculate
   $mean$ and $sd$ for it. Look at the 217th row of your data frame. What
   can you tell about that subject? When was this person tested? Sex? How
   long went this person to school? What score did it have in the
   vocabulary test? Next, extract only those cases from your data frame
   that are male, got tested in 1974 and scored below 2 in the vocabulary
   test.

```{r ex6}
mean(voc$vocabulary)
sd(voc$vocabulary)
voc[217,]
voc[voc$year == "1974" & voc$sex == "Male" & voc$vocabulary < 2,]
```

7. Sort your data frame in a way that people with the same number of years
   of education are grouped together; now, within that, group by their
   score of the vocabulary test. Hint: Use `order()`. Then you can apply
   functions `head()` and `tail()` to look at the first and last 20
   observations, respectively. What do you hypothesize about the
   relationship of "years of education" and "score in vocabulary test?"

```{r ex7}
head(voc[order(voc$education, voc$vocabulary),], n = 20)
tail(voc[order(voc$education, voc$vocabulary),], n = 20)
```

8. We now want to know if there are sex differences in "years of education"
   and "score in vocabulary test." For that we want to look at the mean
   scores for men and women separately. Hint: Use `aggregate()`. You can
   also look at the means for every year checking if, e.\,g., "years of
   education" have increased within the last couple of decades.

```{r ex8}
aggregate(voc[,4:5], by = list(voc$sex), FUN = mean)
aggregate(voc[,4:5], by = list(voc$year), FUN = mean)
```

