```{r, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.show = "")
```

# Simple data analysis in R

R is a free software environment for statistical computing and graphics.
The stats package, which is automatically loaded when R starts up, contains
functions for statistical calculations and random number generation. For a
complete list of functions, see `library(help = stats)`.

## t tests

```{r}
# One-sample t test
rt <- rnorm(100, mean = 400, sd = 20)
t.test(rt, mu = 400)

# Two-sample t test for independent groups
t.test(weight ~ group, PlantGrowth[PlantGrowth$group != "ctrl", ],
       var.equal = TRUE)

# Two-sample t test for dependent groups
t.test(extra ~ group, sleep, paired = T)
```

## Linear models

Let us first fit a simple linear regression.

```{r, fig.dim = c(4, 4.5)}
dat <- data.frame(
  x = c(18, 23, 25, 35, 65, 54, 34, 56, 72, 19, 23, 42, 18, 39, 37),
  y = c(202, 186, 187, 180, 156, 169, 174, 172, 153, 199, 193, 174, 198, 183, 178)
)
lm1 <- lm(y ~ x, dat)
plot(y ~ x, dat)
abline(lm1)
coef(lm1)
predict(lm1)
resid(lm1)
names(lm1)
summary(lm1)
```

```{r, fig.dim = c(8, 8)}
# model diagnostics
par(mfrow = c(2, 2))
plot(lm1)
```


## Analysis of Variance (ANOVA)

An ANOVA is just a special case of a regression where all predictors are
categorical. Using the function `aov()` instead of `lm()` gives us
different results for the extractor functions.

```{r anova}
dat <- read.table(text = "
  id group   score
   1 control  8
   2 control 12
   3 control  7
   4 control 10
   5 control 11
   6 control 12
   7 pro      7
   8 pro      9
   9 pro     15
  10 pro     13
  11 pro     11
  12 pro     16
  13 pro     12
  14 pro      8
  15 pro     13
  16 pro     16
  17 contra   4
  18 contra   5
  19 contra   6
  20 contra   3
  21 contra   8
  22 contra  10
  23 contra   3
  24 contra   9
", header = TRUE)

aov1 <- aov(score ~ group, data = dat)
summary(aov1)
```

We can again look at the model diagnostics.

```{r, fig.dim = c(8, 8)}
# model diagnostics
par(mfrow = c(2, 2))
plot(aov1)
```


# Plotting

There are two graphics systems in R

* Traditional graphics
* Grid graphics

## Traditional plots

```{r, fig.dim = c(4, 4.5)}
plot(cars)
```

Stepwise plotting -- first example:

```{r, fig.dim = c(4, 4.5)}
dat <- read.table(header = TRUE, text = "
  A B rt
  a1 b1 825
  a1 b2 792
  a1 b3 840
  a2 b1 997
  a2 b2 902
  a2 b3 786
", stringsAsFactors = TRUE)

plot(rt ~ as.numeric(B), dat, type = "n", axes = FALSE, xlim = c(.8, 3.2),
  ylim = c(750, 1000), xlab = "Difficulty", ylab = "Mean reaction time (ms)")

# Plot the data points separately for each level of factor A.
points(rt ~ as.numeric(B), dat[dat$A == "a1", ], type = "b", pch = 16)
points(rt ~ as.numeric(B), dat[dat$A == "a2", ], type = "b", pch = 4)

# Add axes and a legend.
axis(side = 1, at = 1:3, expression(B[1], B[2], B[3]))
axis(side = 2)
legend(2.5, 975, expression(A[1], A[2]), pch = c(16, 4), bty = "n",
  title = "Task")
```

Second example:

```{r, fig.dim = c(5, 4.5)}
plot(Sepal.Length ~ Sepal.Width, iris, axes = FALSE, type = "n",
     xlab = "Sepal width", ylab = "Sepal Length")

points(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "setosa"),
      col = "magenta", pch = 21)
points(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "versicolor"),
      col = "red", pch = 22)
points(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "virginica"),
      col = "purple", pch = 23)

axis(1)
axis(2)

legend("topleft", c("setosa", "versicolor", "virginica"),
       pch = 21:23, col = c("magenta", "red", "purple"), bty = "n")

lm1 <- lm(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "setosa"))
lm2 <- lm(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "versicolor"))
lm3 <- lm(Sepal.Length ~ Sepal.Width, subset(iris, iris$Species == "virginica"))

abline(lm1, col = "magenta")
abline(lm2, col = "red")
abline(lm3, col = "purple")
```


## Lattice plots

The lattice package implements Trellis plots in R. These are plots
conditional on other variables. They are perfectly suited for visualizing
complex relationships.

```{r, fig.dim = c(5.3, 5)}
library(lattice)
states <- data.frame(state.x77, state.name = state.name,
                     state.region = state.region) # built-in data sets
xyplot(Murder ~ Population | state.region, states)
```

```{r, fig.dim = c(6, 6), include = FALSE}
# Fine control
xyplot(Murder ~ Population | state.region, states, groups = state.name,
       xlim = c(-5000, 25000), ylab = "Murder rate (per 100000)",

  # panel function
  panel = function(x, y, subscripts, groups){
    panel.grid(-1, -1)      # align with ticks
    panel.abline(lm(y ~ x)) # regression line
    ltext(x, y, labels = groups[subscripts], cex = 1)
  },

  scales = list(tck = .5, cex = .7, x = list(at = seq(0, 20000, 5000))),

  strip = strip.custom(bg = "gray96"), # strip function

  par.strip.text = list(cex = .7)
)
```

```{r, fig.dim = c(6, 3.5)}
xyplot(Sepal.Length ~ Sepal.Width, iris, groups = Species,
       type = c("p", "r"), auto.key = TRUE)
```

Other example to "quickly" look at data:

```{r, fig.dim = c(4, 4), include = FALSE}
xyplot(extra ~ group, sleep, groups = ID, type = "b")
```

```{r, fig.dim = c(5.3, 4)}
xyplot(weight ~ Time | Diet, ChickWeight, groups = Chick,
      type = c("g", "p", "r"))
```


## The Grammar of Graphics `ggplot2`

```{r, include = FALSE}
library(ggplot2)
theme_set(theme_test())
```

R has very powerful graphics function. Creating beautiful (publication
ready) plots might be one of the best reasons to learn R. The R package
`ggplot2` gives you endless possibilities. In recent years it has become
the state of the art way to create plots in R. It has a steeper learning
curve than the functions above (and I recommend to get some knowledge about
them, too, so you understand how plotting in R works), but it is well worth
the effort to invest some time. A good place to start are the following
(online) books:

* Chang, W. (2018). R graphics cookbook: practical recipes for visualizing
  data. O'Reilly Media. https://r-graphics.org/
* Wickham, H. (2016). ggplot2: elegant graphics for data analysis
  Springer-Verlag New York. https://ggplot2-book.org/

The grammar of graphics has a simple structure

```
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(
    mapping = aes(<MAPPINGS>),
    stat = <STAT>,
    position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

```{r}
# load package
library(ggplot2)
# load data
data(mpg)
# ?mpg
mpg
```

### Some examples

```{r, fig.dim = c(3.5, 3.5)}
# start your plot with coordinate system
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +   # then add layers
  theme_bw()                                        # and a theme
```


```{r, fig.dim = c(4, 3.5)}
# even more information
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = fl))
```


What is the difference between these plots?

```{r, fig.dim = c(5, 3.5)}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

```{r, fig.dim = c(4, 3.5)}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

What happens when we do this?

```{r, fig.dim = c(5, 3.5)}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```


### Facets

* Way to visualize additional variables
* Split plot into several facets
* Usually only meaningful for categorical variables
* Used with R's formula notation `y ~ x`

```{r, fig.dim = c(6, 4)}
# one variable
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```


```{r, fig.dim = c(6, 4)}
# two variables
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)
```


### Geometric objects

* A `geom` is the geometrical object that a plot uses to represent data
* For example bar charts, box plots, line charts, ...
* Every `geom` function takes a `mapping` argument `ggplot2` provides over
  40 `geoms`, and extension packages provide even more (see
  https://exts.ggplot2.tidyverse.org/gallery/ for a sampling)

```{r, fig.dim = c(4, 3)}
# scatter plot
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```


```{r, fig.dim = c(4, 3)}
# fitted line
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```


```{r, fig.dim = c(4.5, 3)}
# add another variable
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```



```{r, fig.dim = c(4.5, 3)}
# more layers: points and regression lines
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy, color = drv)) +
  geom_smooth(aes(x = displ, y = hwy, color = drv), method = "lm")
```
