# Exercises Day 2

```{r include = FALSE}
# Hide/show R chunks
#knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.show = "hide", warning = FALSE, message = FALSE)
#setwd("C:/Users/nwickelmaier/Nextcloud/Documents/teaching/iwm/R_intro/exercises/")
rm(list=ls())
```

Write an executable and commented R script.

1. Load the built-in data set \verb+cars+. Find out about the data with
   `?cars`. Create a scatter plot of "Distance" as a function of "Speed."
   Adjust the size of the margins (`mai`) and the spacing between labels
   and axes (`mgp`). Set `type = "n"` and `axes = FALSE` and create the
   plot from scratch, step by step adding `points()`, `axis()`, `legend()`,
   etc.

   Export the plot to a png or pdf file. All labels should be readable, so
   reduce `width` and `height` of the output device.

   ```{r, results = "hide", fig.height = 3.375, fig.width = 3.375}
   par(mai = c(.6,.6,.15,.1), mgp = c(2,.7,0), tck = -.015)
   plot(cars, type = "n", axes = FALSE, xlim = c(0, 30),
     xlab = "Speed (mph)", ylab = "Stopping distance (ft)")
   points(cars, pch = 16)
   axis(side = 1)
   axis(side = 2)
   abline(lm(dist ~ speed, cars), lwd = 2)
   box(bty = "L")
   legend(0, 120, expression(hat(y) == hat(beta)[0] + hat(beta)[1] * Speed),
     lty = 1, lwd = 2, bty = "n")
   ```

2. Load `Vocabulary.txt` into R again. Make a scatter plot of "score in
   vocabulary test" as a function of "years of education." Set `pch = "."`
   and use the `jitter()` function to actually see the structure in the
   data.

   Add the mean vocabulary score for each year of education to your plot.
   Use `aggregate()` or `tapply()` to get these means. Add the standard
   errors (use `arrows()`).

   Export the plot to a file. Play with the graphical parameters. The goal
   is a publication-ready figure.

   ```{r, results = "hide", fig.height = 3.375, fig.width = 3.375}
   voc <- read.table("data/Vocabulary.txt", header = TRUE, stringsAsFactors = TRUE)
   
   # calculate standard errors
   vocm    <- aggregate(vocabulary ~ education, voc, mean)
   vocm$sd <- aggregate(vocabulary ~ education, voc, sd)[,2]
   vocm$n  <- aggregate(vocabulary ~ education, voc, length)[,2]
   vocm$se <- vocm$sd / sqrt(vocm$n)
   
   par(mai = c(.6,.6,.15,.1), mgp = c(2,.7,0), tck = -.015)
   plot(jitter(vocabulary, 2) ~ jitter(education, 2), voc, pch = ".", 
     cex = .8, xlab = "Education (in years)", ylab = "Vocabulary (words correct)",
     col = "darkgrey")
   points(vocabulary ~ I(0:20), vocm, pch = 16)
   with(vocm, arrows(0:20, vocabulary + se, 0:20, vocabulary - se, angle = 90,
     length = .05, code = 3))
   ```

   ```{r, fig.width = 3.375, fig.height = 3.375}
   # Solution with ggplot()
   library(ggplot2)

   ggplot(data = vocm, mapping = aes(x = education, y = vocabulary)) +
     geom_jitter(data = voc, mapping = aes(x = education, y = vocabulary), pch=".",
                 color = "gray") +
     labs(x = "Education (in years)", y = "Vocabulary (words correct)") +
     geom_point() + #geom_line() +
     geom_errorbar(aes(ymin = vocabulary - se, ymax = vocabulary + se), width = 0.5) +
     theme_bw()
   ```

3. Find out if there are sex differences in the relationship of "years of
   education" and "score in vocabulary test," and if these differences
   depend on the year the test was taken in. Load the `lattice` package and
   use the `xyplot()` function. You can also use functions from the
   `ggplot2` package. Hint: Make one scatter plot for each year
   of education with one regression line for men and one for women.

   ```{r, fig.height = 6.75, fig.width = 6.75}
   library(lattice)

   voc$year <- factor(voc$year)

   xyplot(jitter(vocabulary, 2) ~ jitter(education, 2) | year, data = voc,
     groups = sex, type = c("p", "r"), col = c("darkgrey","black"), pch = ".", lwd = 2,
     xlab = "Education (in years)", ylab = "Vocabulary (words correct)", lty = 1:2,
     scales = list(tck = .5, cex = .7),
     par.strip.text = list(cex = .7),
     par.settings = list(strip.background = list(col = "lightgrey")),
     key = list(lines = list(lwd = 2, col = c("darkgrey", "black"), lty = 1:2),
       text = list(levels(voc$sex)), columns = 2)
   )
   ```
   
   ```{r, message = FALSE, fig.width = 7, fig.height = 6.75}
   # Solution with ggplot()
   ggplot(data = voc, mapping = aes(x = education, y = vocabulary, color = sex)) +
     geom_jitter(pch=".") +
     labs(x = "Education (in years)", y = "Vocabulary (words correct)") +
     stat_smooth(method = "lm", se = FALSE) +
     theme_bw() +
     facet_wrap(vars(year))
   ```

4. Subjects are supposed to solve two mental exercises. Time needed will be
   measured in minutes. Between the two exercises subjects get a training
   which is supposed to enhance speed in solving these kind of mental
   exercises. Ten subjects are tested. The following table shows the
   solving times:


   ```{r}
   dat <- data.frame(
     Vp = 1:10,
      A = c(5.68, 4.52, 5.56, 5.17, 4.67, 5.97, 5.65, 6.04, 4.94, 5.17),
      B = c(5.45, 4.37, 5.31, 5.19, 4.18, 5.44, 5.19, 5.84, 4.95, 5.04)
   )
   ```


   ```{r, echo = FALSE, tab.align = "center", results = "show"}
   knitr::kable(dat, col.names = c("Subj", "Exercise A", "Exercise B"), align = "c")
   ```
   a. Create a data frame that contains the depicted variables. Use the
   function `reshape()` to transform the data frame to the "long"
   format.
   
   ```{r}
   dat2 <- reshape(dat, idvar = "Vp", timevar = "Aufg", varying = list(2:3),
                   v.names = "Lzeit", direction = "long")
   dat2$Vp   <- factor(dat2$Vp)
   dat2$Aufg <- factor(dat2$Aufg, labels = c("A", "B"))
   rownames(dat2) <- NULL
   ```

   b. Plot the solving time for each subject depending on the exercise (one
   panel per person). Use `xyplot()` from the `lattice` package. Export
   your graphic as a png or pdf file.

   ```{r, fig.width = 6, fig.height = 5, fig.align = "center"}
   xyplot(Lzeit ~ Aufg | Vp, dat2, as.table = TRUE, type = c("g", "b"),
                pch = 16, xlab = "Exercise", ylab = "Solving time (in min)")
   ```
   
   c. Check if the solving times differ for both exercises using a Wilcoxon
   test ($\alpha = 0.05$).
   
   ```{r}
   with(dat2, wilcox.test(Lzeit[Aufg == "A"], Lzeit[Aufg == "B"], paired = TRUE))
   ```
   
   d. Check the same hypothesis using a $t$ test ($\alpha = 0.05$).
   
   ```{r}
   with(dat2, t.test(Lzeit[Aufg == "A"], Lzeit[Aufg == "B"], paired = TRUE))
   ```
   
5. For $n = 12$ subjects, intelligence (variable $X$) and memory
   performance (variable $Y$) have been measured. The following value pairs
   were obtained:
   
   ```{r}
   dat <- data.frame(
      Vp = 1:12,
     int = c(106, 92, 90, 118, 87, 123, 118, 115, 111, 95, 110, 108),
     mem = c(94, 100, 84, 126, 95, 93, 120, 103, 100, 102, 120, 125)
   )
   ```
   
   
   ```{r, echo = FALSE, tab.align = "center", results = "show"}
   tab <- t(dat[, -1])
   rownames(tab) <- c("$X$", "$Y$")
   knitr::kable(tab, digits = 0, align = "c")
   ```

   a. Create a data frame for these data.

   b. Check the normality assumption separately for each variable using
   quantile-quantile plots. Interpret the results.

   c. Plot memory performance depending on intelligence in a scatter plot
   and add a regression line. Export your graphic as a png or pdf
   file.

   d. Check the hypothesis $\rho_{XY} = 0$ (if $X$ and $Y$ are
   uncorrelated) using an appropriate statistical test ($\alpha = 0.05$).

   ```{r qq_iqmemory, fig.width = 6, fig.height = 3.375}
   par(mfrow = c(1, 2), mai = c(.6, .6, .6, .1), mgp = c(2, .7, 0))
   qqnorm(dat$int, main = "IQ")
   qqline(dat$int)
   qqnorm(dat$mem, main = "Memory")
   qqline(dat$mem)
   ```
   
   ```{r iqmemory, fig.width = 4, fig.height = 4, fig.align = "center"}
   par(mai = c(.6, .6, .1, .1), mgp = c(2, .7, 0))
   plot(mem ~ int, dat, xlab = "IQ", ylab = "Memory")
   abline(lm(mem ~ int, dat))
   cor.test(~ int + mem, dat, method = "pearson")
   ```

