# Advanced topics

The following covers some advanced topics on Programming in R that I usually do
not cover in the first introduction to R. Most of this is very useful when you
want to simulate data in R, e.g., for power simulations.

## Programming resources

Some more advanced resources for R programming are the following:

* Advanced R  https://adv-r.hadley.nz/

* Happy Git and GitHub for the useR https://happygitwithr.com/

* R Programming for Research https://geanders.github.io/RProgrammingForResearch/

* Building reproducible analytical pipelines with R https://raps-with-r.dev/

* Data Skills for Reproducible Science https://psyteachr.github.io/msc-data-skills/

---
nocite: |
  @Venables2022, @Ligges2008, @Wickham_styleguide, @Anderson2023
---

## Writing Functions

* Functions in R consist of
  - a name
  - a pair of brackets
  - the arguments (none, one, or more)
  - a return value (visible, invisible, \texttt{NULL})

* Arguments are passed
  - either without name (in the defined order)

    -> positional matching
 
  - or with name (in arbitrary order)

    -> keyword matching

* Even if no arguments are passed, the brackets need to be written, e.g.,
  `ls()`, `dir()`, `getwd()`

* Entering only the name of a function without brackets will display the R code
  of that function

### Implementing a function for a simple two-sample t test

Let us implement a simple function ourselves. A function that implements a
two-sample t test

\begin{equation*}
  T = \frac{\bar{x} - \bar{y}}
      {\sqrt{\hat{\sigma}^2\, \left(\frac{1}{n} + \frac{1}{m}\right)}}
    = \frac{\bar{x} - \bar{y}}
      {\sqrt{\frac{(n-1) \, s_x^2 + (m-1) \, s_y^2}
          {n+m-2}\cdot \left(\frac{1}{n} + \frac{1}{m}\right)}}
\end{equation*}
with
\begin{equation*}
  T \sim t(n+m-2)
\end{equation*}

```{r}
# Example: a handmade t test function
twosam <- function(y1, y2){              # definition
  n1  <- length(y1); n2  <- length(y2)   # body
  yb1 <- mean(y1);   yb2 <- mean(y2)
  s1  <- var(y1);    s2  <- var(y2)
  s   <- ((n1 - 1)*s1 + (n2 - 1)*s2)/(n1 + n2 - 2)
  tst <- (yb1 - yb2)/sqrt(s*(1/n1 + 1/n2))
  tst                                    # return value, can also be a list
}

# Calling the function
tstat <- twosam(PlantGrowth$weight[PlantGrowth$group == "ctrl"],
                PlantGrowth$weight[PlantGrowth$group == "trt1"])
tstat
```
### Named arguments and defaults

If there is a function \texttt{fun1} defined by

```{r, eval = FALSE}
fun1 <- function(data, data.frame, graph, limit) { 
...
}
```
then the function may be invoked in several ways, for example

```{r, eval = FALSE}
fun1(d, df, TRUE, 20)
fun1(d, df, graph = TRUE, limit = 20)
fun1(data = d, limit = 20, graph = TRUE, data.frame = df)
```

All of them are equivalent (cf. positional matching and keyword matching)

In many cases, arguments can be given commonly appropriate default values, in
which case they may be omitted altogether from the call

```{r, eval = FALSE}
fun1 <- function(data, data.frame, graph = TRUE, limit = 20) {
...
}
```

It could be called as

```{r, eval = FALSE}
ans <- fun1(d, df)
```

which is now equivalent to the three cases above, or as

```{r, eval = FALSE}
ans <- fun1(d, df, limit = 10)
```

which changes one of the defaults.

### Exercise

Write a function in R that cumulatively sums up the values of vector $\mathbf{x}
= (1~2~3~4 \dots 20)'$. The result should look like: 
$\mathbf{y} = (1~3~6~10 \dots 210)'$.

```{r}
# TODO: Add solution to exercise
```

## Conditional execution

When programming, a distinction of cases is often necessary for

* checking of arguments

* return of error messages

* interrupting a running process

* case distinction, e.\,g., in mathematical expressions

Conditional execution of code is available in R via

```{r, eval = FALSE}
if(expr_1) {
  expr_2
} else {
  expr_3
}
```

where `expr_1` must evaluate to a single logical value.

```{r}
x <- 5

# Example 1
if(!is.na(x)) y <- x^2 else stop("x is missing")

# Example 2
if(x == 5){   # in case x = 5:
  x <- x + 1  # add 1 to x and
  y <- 3      # set y to three
} else        # else:
  y <- 7      # set y to seven

# Example 3
if(x < 99) cat("x is smaller than 99\n")

## Vectorized version with ifelse() function

# Example 1
ifelse(x == c(5, 6), c("A1", "A2"), c("A3", "A4"))

# Example 2
x <- -2:2
ifelse(x < 0, -x, x)
```

### Exercise

Implement the following function in R:

\[
f(x) =
\begin{cases}
 -1 & \text{if } x < 0,\\
  0 & \text{if } x = 0,\\
  1 & \text{if } x > 0.
\end{cases}
\]

```{r}
# TODO: Add solution to exercise
```

## Loops

* Loops are necessary to execute repeating commands

* Especially for simulations, loops are often used

* In this case, the same functions or commands are executed for different random numbers

* There are `for()` and `while()` loops for repeated execution

* The most simple "loop" is `replicate()`

```{r}
# Example for central limit theorem

y <- runif(100)     # Draw random numbers
hist(y)

x <- replicate(1000, mean(runif(100)))
hist(x)

# Examples repeat and while

i <- 0
repeat {
  i <- i + 1              # Add 1 to i
  if (i == 3) break       # Stop if i = 3
}

while (i > 1) {
  i <- i - 1              # As long as i > 1, subtract 1
}

# Example vector operations

x <- c(3, 6, 4, 8, 0)     # Vector of length 5

for(i in x)
  print(sqrt(i))

for(i in seq_along(x))    # Same using indices
  print(sqrt(x[i]))

for(i in seq_along(x)){   # For all i [1,2,3,4,5]'
  x[i] <- x[i]^2          # square ith element of x
  print(x[i])             # and show it on monitor
}

x^2                       # BETTER
```

### Exercise

* Create a vector $\mathbf{x} = (3~5~7~9~11~13~15~17)'$ with a for-loop

  Tip: Use the formula $n\cdot2 + 1$

* Implement two different methods:

  1. Allocate memory: Start with a vector of zeros and the correct length and
     replace its elements iteratively
    
  2. Growing: Start with a `NULL` object and iteratively add new results

Tip: The first method is more efficient, especially for long vectors.

```{r}
# TODO: Add solution to exercise
```

## Avoiding loops

The `apply()` family of functions may be used in many places where in
traditional languages loops are employed. Using vector based alternatives is
usually much faster in R.

* Matrices and arrays: `apply()`

* Data frames, lists and vectors: `lapply()` and `sapply()`

* Group-wise calculations: `tapply()`

```{r}
# Example apply()

X <- matrix(c(4, 7, 3, 8, 9, 2, 5, 6, 2, 3, 2, 4), nrow = 3, ncol = 4)

# Calculate row maxima
res <- numeric(nrow(X))
for(i in 1:nrow(X)){
  res[i] <- max(X[i,])
}

# or:
apply(X, 1, max)  # Maximum for each row
apply(X, 2, max)  # Maximum for each column

# Example lapply()
L <- list(x = 1:10, y = 1:5 + 0i)
lapply(L, mean)       # Keep list with data type
sapply(L, mean)       # Create vector, same data type

sapply(iris, class)   # Work column wise on data frame

# Example tapply()
data(Oats, package = "nlme")
with(Oats, tapply(yield, list(Block, Variety), mean))

data(warpbreaks)
tapply(warpbreaks$breaks, warpbreaks$tension, sum)
tapply(warpbreaks$breaks, warpbreaks[ , -1], mean)
```

### Exercise

* Load the iris data set into R using `data(iris)`

* Write a for-loop to calculate the means for the dependent variables
  (columns 1 to 4)

* Think of as many vector based alternatives as possible to avoid this loop and
  calculate the column means


```{r}
# TODO: Add solution to exercise
```

## Random number generation

Most distributions that R handles have four functions. There is a root name,
e.g., the root name for the normal distribution is `norm`. This root is prefixed
by one of the letters `p`, `q`, `d`, `r`.

|      |                                                             |
| ---- | ----------------------------------------------------------- |
| `p`  | probability: the cumulative distribution function (CDF)     |
| `q`  | quantile: the inverse CDF                                   |
| `d`  | density: the probability (density) function (PDF)           |
| `r`  | random: a random variable having the specified distribution |

See `?Distributions` for a list of distributions or the CRAN task view
https://cran.r-project.org/view=Distributions. The random number generator in R
is _seeded_: Upon restart of R, new random numbers are generated. To replicate
the results of a simulation, the seed (starting value) can be set explicitly
with `set.seed()`.

```{r}
# Examples
rnorm(10)     # Draw from standard normal distribution
rpois(10, 1)  # Draw from Poisson distribution

# Sampling with or without replacement from a vector
sample(1:5, size = 10, replace = TRUE)

# Set seed
set.seed(1223)  # On each run, random numbers will be identical
runif(3)
```

### Exercise

* Create a data frame with four variables and 20 observations:

  1. $X \sim N(\mu=100, \sigma^2=15^2)$
  2. $Y \sim Bin(n=10, p=0.2)$
  3. $Z \sim Pois(\lambda=1)$
  4. $S = X + Y + Z$

```{r}
# TODO: Add solution to exercise
```



